---
title: "Decision Tree Statistical Analysis"
author: "Sara Benist"
date: "2023-04-06"
output: html_document
---

# Processing script with R script code

This Quarto file runs the statistical analysis code from the `statistical_analysis.r` script. The best way to view the output of the script is to render the qmd and view the html document. The processed data is used for model fitting.

*Note: this file takes approximately 10 minutes to render*

To start, the needed packages are loaded, and the `processeddata.rda` is loaded from the `processed_data` folder. 

# Setup
The `statistical_analysis.r` script needs to be indicated as the file from which to pull code chunks from. 
```{r, cache=FALSE}
knitr::read_chunk('statistical_analysis.R')
```

# Load packages and data
The packages and data are loaded.
```{r, packages, message = FALSE, warning = FALSE}
```

```{r, loaddata}
```

# Basic modelling
For some basic modeling, I will be fitting linear models to the TB outcomes with the subgroups as predictors. Subgroups are the main difference between populations other than country which is why I chose these as the bivariate models. This will offer another view of the data. The prediction model also offers a way to explore the data by submitting a sample prediction of TB incidence for females and males in Central African Republic.
```{r, basicmodels}
```
The models indicate that subgroup has a significant influence on the outcomes since the estimates for the subgroup (being Male primarily or living in rural areas) increases the incidence, mortality, or prevalence. This is expected from the exploratory analysis. 

The biggest issue with the linear modeling is the assumption that the relationship between predictors and outcomes is linear, and the model cannot be tuned and replicated.

# Model fitting
Now that the data has been explored, I would like to start looking at models for the main outcomes and predictors of the data set. First, we will split the data into testing and training sets and set up the cross-validation folds. The training set will be used to tune the model parameters and the testing set will be used to test model performance on new data.

```{r, splitdata}
```

I will be tuning decision tree models for the main outcomes of TB incidence, TB mortality, and TB prevalence for both high burden countries and the other burden set. In the `boosted_statistical_analysis.qmd` file, a boosted tree model for each outcome is created for comparison. In order to be clear when saving the resulting tables and figures, the other burden outcomes from the decision tree model are labeled as `obdt` and the high burden as `hbdt`. The testing data will be fit to the best performing models. 

## Decision tree model for TB Incidence
```{r, obdtincidencemodel}
```
The best tree model for TB incidence using the other burden countries has a depth of 4 with an RMSE of 51.6. This is a very low level of performance in predicting TB incidence (ideally would be closer to 1). The fitted model to the training data has a RMSE of 29.8.

```{r, hbdtincidencemodel}
```
The best tree model for TB incidence using the high burden countries has a depth of 4 with an RMSE of 159. This is also a low level of performance in predicting TB incidence. The fitted model to the training data has a RMSE of 81.3 which was much higher compared to the full dataset. 

## Decision tree model for TB Mortality
```{r, obdtmortalitymodel}
```
The best tree model for TB mortality using the other burden has a depth of 8 with an RMSE of 8.85. The fitted model to the training data has a RMSE of 7.21.

```{r, hbdtmortalitymodel}
```
The best tree model for TB mortality using the high burden countries has a depth of 4 with an RMSE of 20.4. The fitted model to the training data has a RMSE of 10.3.

## Decision tree model for TB Prevalence

```{r, obdtprevalencemodel}
```
The best tree model for TB prevalence using the other burden countries has a depth of 1 with an RMSE of 294. This is a very low level of performance in predicting TB prevalence. The fitted model to the training data has a RMSE of 240.

```{r, hbdtprevalencemodel}
```
The best tree model for TB prevalence using the high burden countries has a depth of 1 with an RMSE of 322. This is a very low level of performance in predicting TB incidence. The fitted model to the training data has a RMSE of 254.

The models were overall not a good predictor of TB outcomes. This is possibly due to the large variance in the outcomes as well as the summarizing required to de-aggregate the data and predict outcomes. Country was overwhelmingly the most important predictor.

## Summarise models
Overall, the models were not great predictors of TB outcomes. The mortality performed the best, and all high burden models had lower RMSE values by about half of the other burden models. 
In the `boosted_statistical_analyisis.qmd` file, I created boosted tree models as a comparison. The boosted tree models should perform better than a decision tree model. 
```{r, summary}
```
