---
title: "Data Exploring, Cleaning, and Wrangling"
author: "Sara Benist"
date: "2023-02-24"
output: html_document
---

# Processing script with code pulled in

This is Quarto file processes code chunks located in the `processingcode.r` script.

# Setup

The script below indicates which R file the Quarto should pull from.

```{r, include=FALSE, cache=FALSE}
knitr::read_chunk('processingcode.R')
```

Load needed packages for data cleaning.

```{r, packages, message = FALSE, warning = FALSE}
```

# Data loading

Load data into `raw_data` folder and assign to `rawdata`

```{r, loaddata}
```

# Explore data

Let's look at the data using `glimpse()` and `skim()` which both present an overview of the data. Running `gg_miss_var()` will check for missing data by each variable.

```{r, exploredata1}
```

The data has the countries listed under setting, year of the data, the indicator that is being studied, and the dimension and subgroup for each observation. The estimate column holds the actual percentage or number of interest. Other variables on the data set contain formatting information.

The `skim()` function provides more information of the missing data and the numeric summaries (mean, median, percentiles, etc). The `gg_miss_var()` helps visually display the missing data.

I would also like a list of the indicator names and a count of how many countries are in the data set.

```{r, exploredata2}
```

There are 14 indicators and 194 countries.

# Cleaning

Now the data cleaning can get started. This phase will coincide with data exploration, so more cleaning will most likely occur in the `exploratory_analysis.qmd`.

To start, I want the indicators to move to columns using `pivot_wider()` function so each indicator has a column. I will also subset the columns I would like to keep moving into the exploratory phase. The wide version of the data will be assigned to `wide_data`.

I will also be looking specifically at high TB burden countries. The high burden countries list is updated by the WHO and can be found [here](https://www.who.int/news/item/17-06-2021-who-releases-new-global-lists-of-high-burden-countries-for-tb-hiv-associated-tb-and-drug-resistant-tb). I will subset these countries in the `highburden` object. The countries with low or mid level burden will be sorted into `otherburden`.

```{r, cleandata1}
```

The `wide_data` had 7473 rows and 20 columns with 194 countries. The `highburden` object has 3761 rows and 20 columns with 49 countries. The `otherburden` object has 3712 rows and 20 columns with 145 countries. 


## Subset by indicator

Next, I would like to check each indicator for potential data cleaning issues I need to address. I will use the `filter()` and `select()` functions to subset each indicator and `summary()` to check each indicator.

Here, the BCG coverage, drug resistance, and case detection rate indicators are checked.

```{r, cleandata2}
```

Here, the percentage reporting TB is spread through cough, overall and by sex, indicators are checked.

```{r, cleandata3}
```

Here, the TB incidence and TB mortality indicators are checked.

```{r, cleandata4}
```

Here, the percentage wanting to keep TB infection a secret, overall and by sex, indicators are checked.

```{r, cleandata5}
```

Here, the TB prevalence and prevalence to notification ratio indicators are checked.

```{r, cleandata6}
```

Finally, the percentage of affected people suffering catastrophic costs due to TB indicator is checked.

```{r, cleandata7}
```

Most of the data seems to be from 2004-2020, and the estimates are within the bounds of the indicator (such as all the percentages are between 0 and 100).

## Missing data

While examining the code using the summary tables, a few of the indicators have missing data.

Looking over the data more closely, the missing data (marked as `NA`) is due to a country and year having data for only one subgroup in the dimension. Using the `drop_na()` function would drop all of the data since every row is incomplete. For now, I won't change the data until I have explored more. 

I would also like to mention the odd population minimums for BCG coverage, drug resistance, and case detection rate which are 18.9, 26, and 18 respectively. These seem like odd population counts when the maximum is in the 600,000s. I believe these numbers are due to the subsetting of the indicators by dimension subgroups, so I will not be changing these numbers.

# Modelling setup

The final analysis will be predicting tuberculosis outcomes from the indicators. Due to the nature of the data, each indicator must be summarized for males and females (aggregate data can measure males or someone in the lowest economic quintile but cannot predict the indicator measure for males in the lowest economic quintile). To set up the data for modelling, the data is summarized by sex and indicator for each country in the high burden and other burden data sets. The incidence and mortality outcomes consist of indicators with sex as a dimension, and the prevalence outcome consists of rural or urban subgroups.

```{r, modelsetup}
```

# Save data

The data is (mostly) finished cleaning and can be saved. Since I have separated the indicators, we will save each object into the `processed_data` file under `processeddata.rda`

The data is saved as an rda file rather than an RDS file in order to save each item individually. This file can be loaded into the `exploratory_analysis.qmd` for the next step.

```{r, savedata}
```
